An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-18 thesis/compact-finite-difference.tex
fix typos

Here, we discuss the application of the
tridiagonal solver developed in the previous section
to the evaluation of compact finite differences---which
are used widely in the direct numerical simulation
of fluid flows.
%
\begin{figure}
\begin{center}
\includegraphics[height=150pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}
%
In typical DNS applications,
the computational domain considered
is a 3-dimensional,
regular, Cartesian grid with
$nx$, $ny$ and $nz$ grid points and grid spacing of
$dx$, $dy$ and $dz$ in the
respective coordinate directions
(Fig. \ref{fig:computational-domain}).
Structured grids such as these
are widely used because they lend themselves naturally
to finite-difference methods.
To resolve all the spatial features of the flow,
the grid spacing is kept very small.
Consequently,
to simulate flows of practical sizes,
the number of grid points must be very large.
Because of the high computational cost
and memory requirements associated with
the large number of grid points
parallelism becomes mandatory.
We have spoken so far of GPUs as massively parallel systems.
However, each GPU has a very limited amount of global memory:
current NVIDIA Tesla K20 GPUs have about 4 Gigabytes.
Thus, to accomodate the
larger problem sizes in DNS,
the use of \emph{multiple} GPUs is necessary,
which introduces a second level of parallelism.
In the next section,
we describe strategies to distribute
the problem domain among the GPUs
in such dual-level parallel systems.

@@

\begin{figure}
\begin{center}
\includegraphics[width=200pt]{img/compact-distributed-restricted.eps}
\caption{Compact finite differences - distributed GPUs and
    restriced in one direction}
\label{fig:compact-distributed-restricted}
\end{center}
\end{figure}

For larger problems,
a distributed system may be required.
Here, the simplest strategy is to use a domain decomposition
as shown in \ref{fig:compact-distributed-restricted}.
Here again, 
no inter-GPU communication is required.
By restricting the distribution along one coordinate direction,
the ease of solution of the tridaigonal systems is maintained.

@@

\item For domains that are much longer along one coordinate
    direction compared to the other(s),
    the subdomains may become impractically \emph{slender}.
\end{enumerate}
%
Such decomposition strategies are therefore, generally applicable
when compact finite difference schemes are used only
in a single direction.
For the other coordinate directions,
explicit finite difference schemes may be used,
which have the disadvantages discussed earlier.
To accomodate compact finite difference schemes in all the coordinate
directions,
the domain decomposition generally must be performed in all directions,
as shown in Fig. \ref{fig:compact-distributed-all}.
In this strategy,
the grid lines in all coordinate directions
are interrupted by the subdomain boundaries.
Thus,
inter-GPU communication is required.
For the right hand sides evaluation,
each GPU must communicate information at the subdomain boundaries
with neighbouring GPUs.
For example, in Fig. \ref{fig:compact-distributed-all},
the right-most grid points in the subdomain of
GPU 2 require data from the
left-most grid points in the subdomain of
GPU 6 when evaluating derivatives in the horizontal direction.
Similarly,
the left-most grid points in the subdomain of
GPU 6 require data from the
right-most grid points in the subdomain of
GPU 2.
Thus, a ``swapping'' of the boundary information
is required at each of the subdomain boundaries.

@@

The algorithm described in
Sec. \ref{subsec:distributed-tridiagonal-algorithm}
is applicable to a single tridiagonal system distributed
accross several processors.
When solving a distributed tridiagonal system
for several distributed grid lines,
we note that
the secondary systems
[Eq. (\ref{eqn:secondary-system-1-expanded}) and
 Eq. (\ref{eqn:secondary-system-2-expanded})]
are dependent only on $\bm{a}$, $\bm{b}$ and $\bm{c}$,
the tridiagonal coefficients of the global tridiagonal system.
These are the same for all local grid lines,
so they are solved for a single local grid line in each subdomain.
Only the system in Eq. (\ref{eqn:primary-system-expanded})
is solved for all the grid lines in a subdomain.
For a single tridiagonal system,
the reduced system requires \emph{global} communication
of the values $\{u^p_1, u^p_m\, l^p_1, l^p_m\, x^p_{r,1} x^p_{r,m}\}$
from each process $p$.
When solving for several grid lines,
the left hand side may be assembled for a single grid line.
The right hand sides, however, must be assembled for every grid line,
and the system is solved for each right hand side.

@@

Our algorithm is outlined in \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contigiously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
Python library.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

>>>>

Here, we discuss the application of the
tridiagonal solver developed in the previous section
to the evaluation of compact finite differences---which
are used widely in the direct numerical simulation
of fluid flows.
%
\begin{figure}
\begin{center}
\includegraphics[height=150pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D.}
\label{fig:computational-domain}
\end{center}
\end{figure}
%
In typical DNS applications,
the computational domain considered
is a 3-dimensional,
regular, Cartesian grid with
$nx$, $ny$ and $nz$ grid points and grid spacing of
$dx$, $dy$ and $dz$ in the
respective coordinate directions
(Fig. \ref{fig:computational-domain}).
Structured grids such as these
are widely used because they lend themselves naturally
to finite-difference methods.
To resolve all the spatial features of the flow,
the grid spacing is kept very small.
Consequently,
to simulate flows of practical sizes,
the number of grid points must be very large.
Because of the high computational cost
and memory requirements associated with
the large number of grid points
parallelism becomes mandatory.
We have spoken so far of GPUs as massively parallel systems.
However, each GPU has a very limited amount of global memory:
current NVIDIA Tesla K20 GPUs have about 4 Gigabytes.
Thus, to accommodate the
larger problem sizes in DNS,
the use of \emph{multiple} GPUs is necessary,
which introduces a second level of parallelism.
In the next section,
we describe strategies to distribute
the problem domain among the GPUs
in such dual-level parallel systems.

@@

\begin{figure}
\begin{center}
\includegraphics[width=200pt]{img/compact-distributed-restricted.eps}
\caption{Compact finite differences - distributed GPUs and
    restricted in one direction}
\label{fig:compact-distributed-restricted}
\end{center}
\end{figure}

For larger problems,
a distributed system may be required.
Here, the simplest strategy is to use a domain decomposition
as shown in \ref{fig:compact-distributed-restricted}.
Here again, 
no inter-GPU communication is required.
By restricting the distribution along one coordinate direction,
the ease of solution of the tridiagonal systems is maintained.

@@

\item For domains that are much longer along one coordinate
    direction compared to the other(s),
    the subdomains may become impractically \emph{slender}.
\end{enumerate}
%
Such decomposition strategies are therefore, generally applicable
when compact finite difference schemes are used only
in a single direction.
For the other coordinate directions,
explicit finite difference schemes may be used,
which have the disadvantages discussed earlier.
To accommodate compact finite difference schemes in all the coordinate
directions,
the domain decomposition generally must be performed in all directions,
as shown in Fig. \ref{fig:compact-distributed-all}.
In this strategy,
the grid lines in all coordinate directions
are interrupted by the subdomain boundaries.
Thus,
inter-GPU communication is required.
For the right hand sides evaluation,
each GPU must communicate information at the subdomain boundaries
with neighbouring GPUs.
For example, in Fig. \ref{fig:compact-distributed-all},
the right-most grid points in the subdomain of
GPU 2 require data from the
left-most grid points in the subdomain of
GPU 6 when evaluating derivatives in the horizontal direction.
Similarly,
the left-most grid points in the subdomain of
GPU 6 require data from the
right-most grid points in the subdomain of
GPU 2.
Thus, a ``swapping'' of the boundary information
is required at each of the subdomain boundaries.

@@

The algorithm described in
Sec. \ref{subsec:distributed-tridiagonal-algorithm}
is applicable to a single tridiagonal system distributed
across several processors.
When solving a distributed tridiagonal system
for several distributed grid lines,
we note that
the secondary systems
[Eq. (\ref{eqn:secondary-system-1-expanded}) and
 Eq. (\ref{eqn:secondary-system-2-expanded})]
are dependent only on $\bm{a}$, $\bm{b}$ and $\bm{c}$,
the tridiagonal coefficients of the global tridiagonal system.
These are the same for all local grid lines,
so they are solved for a single local grid line in each subdomain.
Only the system in Eq. (\ref{eqn:primary-system-expanded})
is solved for all the grid lines in a subdomain.
For a single tridiagonal system,
the reduced system requires \emph{global} communication
of the values $\{u^p_1, u^p_m\, l^p_1, l^p_m\, x^p_{r,1} x^p_{r,m}\}$
from each process $p$.
When solving for several grid lines,
the left hand side may be assembled for a single grid line.
The right hand sides, however, must be assembled for every grid line,
and the system is solved for each right hand side.

@@

Our algorithm is outlined in \ref{fig:compact-algorithm},
and is described for the case of evaluating derivatives
in the direction along which successive
function values in a subdomain are stored contiguously in memory,
i.e., the ``fastest'' coordinate direction.
Each step of the algorithm must be implemented on the GPU
to avoid data transfer to and from the CPU,
which is prohibitively expensive for large problems.
Thus, we have several kernels to implement the algorithm.
For communication of data between processes,
we use the Message Passing Interface (MPI),
and leverage the NVIDIA GPUDirect Technology
for GPU-GPU communication.
MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
Python library.
The purpose of each CUDA kernel and MPI call used
is described in Table \ref{table:compact-algorithm}.

### 2015-11-18 thesis/introduction-and-background.tex
fix typos

The workhorse for computation in the above described
parallel systems is the CPU core,
however, more recently,
Graphics Processing Units (GPUs)
are being used for performing intensive calculations.
GPUs, while themselves being highly parallel processors,
can also function as accelerators in distributed memory systems
(GPU clusters).
However,
applications that exploit such systems
need careful redesign of algorithms---and
sometimes, substantial changes to code---to
see significant performance improvements.
This is because the CPU and GPU have very different architectures,
and any na\"{\i}ve ``parallelization'' of algorithms
designed for the CPU
is likely not take full advantage of the GPU's
memory heirarchy and compute ability.

@@

The ability of a finite difference scheme
to accomodate large wavenumbers
is extremely relevant in
computational fluid dynamics applications
~\cite{KravchenkoEffNumErr}.
In DNS, the scheme
\emph{must} capture the
rapidly varying characteristics of the flow
associated with the turbulence.
For this purpose,
higher order explicit schemes
may be considered.
While they are straightforward to compute,
they are associated with large stencil widths.
In most applications,
large stencil widths are undesirable.
This is because
the arithmetic intensity increases with stencil size, i.e.,
a larger number of computations must be performed per grid point.
Further,
the amount of boundary information that must be exchanged
between parallel processes increases,
which can seriously affect overall performance.

@@

The CUDA application programming interface (API)
is availabe as an extension to
a programming languages (C, C++, Fortran, etc.),
allowing certain portions of code to execute on the GPU.
The rest of the code is executed as usual on the CPU.
In CUDA terminology,
the CPU is referred to as the \emph{host},
and the GPU is referred to as the \emph{device}.
The special pieces of code
that execute on the GPU are known as \emph{kernels}.
Kernels have similar syntax
as ``host code'',
but in general,
are more restricted in the
features of the underlying language they can use.
In C, for example, kernels are written as \emph{functions},
and are called by the host code
using (almost) the same conventions.
Thus, a C program that uses the GPU
looks and behaves very much like a normal C program,
but includes special calls to these kernels.
When the application is launched,
a CPU thread executes the host-code as usual,
but upon encountering a call to the kernel,
it passes program control over to the GPU.
After the kernel is finished executing,
control is passed back to the CPU,
and this process may repeat when another kernel call is encountered.
The CPU and GPU may also operate \emph{asynchronously},
i.e.,
control may be passed back to the CPU \emph{before} kernel completion,
in which case
explicit synchronization between the host and device may be necessary.
Apart from kernel calls, the host code can also call functions to
allocate and deallocate device memory,
query device information,
perform device synchronization, etc.

@@

From the hardware perspective,
an NVIDIA GPU may be viewed primarily as
a collection of so-called Streaming Microprocessors (SMs).
When a kernel is launched with specified
grid size (number of blocks)
and block size (threads per block),
each block is assigned to an SM.
Threads within a thread block can execute concurrently on the SM,
and an SM can execute several thread blocks concurrently.
It is imperative that blocks are able to run
independently, in series or in parallel.
This feature gives CUDA programs
their scalability \ref{fig:gpu-scaling}.
The same program runs faster on a GPU with a larger number
of SMs, simply because more blocks may run concurrently.
The number of thread blocks that an
SM can execute concurrently
is limited by a number of factors,
and maximizing this number is often
key to obtaining good performance.
Another key consideration in the implementation
of algorithms on GPUs is the
available \emph{memory heirarchy}.
Threads executing a kernel can read and write
to several different memory spaces,
each with different capacity, latency and bandwidth.

@@

Shared memory is organized as a collection of  4 byte ``words''.
Each consecutive word in shared memory belongs
to a different bank - modern GPUs have 32 banks.
The GPU schedules
thread execution within a block
in groups of 32 (termed as thread \emph{warps}).
Ideally, these 32 threads
can access shared memory locations concurrently.
However, when two or more threads in a warp access
words from the same bank,
a \emph{bank conflict} is said to occurr,
and the access is serialized.
Thus,
a warp of 32 threads reading
32 successive \texttt{float}s (4 bytes) in a shared memory array
is perfectly parallelized.
But a warp of 32 threads reading
32 alternating \texttt{float}s in a shared memory array
leads to bank conflicts (Fig. \ref{fig:bank-conflicts}).
Because the floats are placed in consecutive banks,
the 1st and 17th threads will acess floats from the same bank,
as will the 2nd and 18th threads, and so on.
This specific kind of bank conflict is termed a
\emph{two-way} bank conflict -
threads access two words from the same bank.
It is observed that higher-order bank conflicts may occur,
if threads in a warp access
\texttt{float}s in strides of 4,
a \emph{four-way} bank conflict can occur.
In the worst case, successive threads may access
\texttt{float}s in strides of 32,
in which case all threads in the warp
request memory from the same bank:
a emph{32-way} bank conflict.

@@

While shared memory and registers can service memory requests
much faster than global memory,
their overuse can lead to performance degradation.
The amount of shared memory per SM
and the numer of registers per SM is limited.
For the current NVIDIA Tesla K20 accelerator,
the amount of shared memory per SM is limited to 48 KiB,
and the number of registers per SM is limited to 65536.
These are also the limits
on the resources that can be allocated for each \emph{block}.
However,
allocating 48 KiB shared memory or using 65536 registers 
for each block is ill-advised,
as this effectively restricts the number of blocks that
each SM can run at any given time to 1.
If each block allocates 24 KiB of shared memory,
then an SM can run only 2 blocks concurrently.
Thus, the resources allocated per-block affects
the overall parallelism that can be exploited from the GPU.

@@

    \item The limited amount of shared memory places restricions
        on the size of tridiagonal systems that can be solved,
        and also on the number of systems that can be solved
        concurrently.
\end{enumerate}

>>>>

The workhorse for computation in the above described
parallel systems is the CPU core,
however, more recently,
Graphics Processing Units (GPUs)
are being used for performing intensive calculations.
GPUs, while themselves being highly parallel processors,
can also function as accelerators in distributed memory systems
(GPU clusters).
However,
applications that exploit such systems
need careful redesign of algorithms---and
sometimes, substantial changes to code---to
see significant performance improvements.
This is because the CPU and GPU have very different architectures,
and any na\"{\i}ve ``parallelization'' of algorithms
designed for the CPU
is likely not take full advantage of the GPU's
memory hierarchy and compute ability.

@@

The ability of a finite difference scheme
to accommodate large wavenumbers
is extremely relevant in
computational fluid dynamics applications
~\cite{KravchenkoEffNumErr}.
In DNS, the scheme
\emph{must} capture the
rapidly varying characteristics of the flow
associated with the turbulence.
For this purpose,
higher order explicit schemes
may be considered.
While they are straightforward to compute,
they are associated with large stencil widths.
In most applications,
large stencil widths are undesirable.
This is because
the arithmetic intensity increases with stencil size, i.e.,
a larger number of computations must be performed per grid point.
Further,
the amount of boundary information that must be exchanged
between parallel processes increases,
which can seriously affect overall performance.

@@

The CUDA application programming interface (API)
is available as an extension to
a programming languages (C, C++, Fortran, etc.),
allowing certain portions of code to execute on the GPU.
The rest of the code is executed as usual on the CPU.
In CUDA terminology,
the CPU is referred to as the \emph{host},
and the GPU is referred to as the \emph{device}.
The special pieces of code
that execute on the GPU are known as \emph{kernels}.
Kernels have similar syntax
as ``host code'',
but in general,
are more restricted in the
features of the underlying language they can use.
In C, for example, kernels are written as \emph{functions},
and are called by the host code
using (almost) the same conventions.
Thus, a C program that uses the GPU
looks and behaves very much like a normal C program,
but includes special calls to these kernels.
When the application is launched,
a CPU thread executes the host-code as usual,
but upon encountering a call to the kernel,
it passes program control over to the GPU.
After the kernel is finished executing,
control is passed back to the CPU,
and this process may repeat when another kernel call is encountered.
The CPU and GPU may also operate \emph{asynchronously},
i.e.,
control may be passed back to the CPU \emph{before} kernel completion,
in which case
explicit synchronization between the host and device may be necessary.
Apart from kernel calls, the host code can also call functions to
allocate and deallocate device memory,
query device information,
perform device synchronization, etc.

@@

From the hardware perspective,
an NVIDIA GPU may be viewed primarily as
a collection of so-called Streaming Microprocessors (SMs).
When a kernel is launched with specified
grid size (number of blocks)
and block size (threads per block),
each block is assigned to an SM.
Threads within a thread block can execute concurrently on the SM,
and an SM can execute several thread blocks concurrently.
It is imperative that blocks are able to run
independently, in series or in parallel.
This feature gives CUDA programs
their scalability \ref{fig:gpu-scaling}.
The same program runs faster on a GPU with a larger number
of SMs, simply because more blocks may run concurrently.
The number of thread blocks that an
SM can execute concurrently
is limited by a number of factors,
and maximizing this number is often
key to obtaining good performance.
Another key consideration in the implementation
of algorithms on GPUs is the
available \emph{memory hierarchy}.
Threads executing a kernel can read and write
to several different memory spaces,
each with different capacity, latency and bandwidth.

@@

Shared memory is organized as a collection of  4 byte ``words''.
Each consecutive word in shared memory belongs
to a different bank - modern GPUs have 32 banks.
The GPU schedules
thread execution within a block
in groups of 32 (termed as thread \emph{warps}).
Ideally, these 32 threads
can access shared memory locations concurrently.
However, when two or more threads in a warp access
words from the same bank,
a \emph{bank conflict} is said to occur,
and the access is serialized.
Thus,
a warp of 32 threads reading
32 successive \texttt{float}s (4 bytes) in a shared memory array
is perfectly parallelized.
But a warp of 32 threads reading
32 alternating \texttt{float}s in a shared memory array
leads to bank conflicts (Fig. \ref{fig:bank-conflicts}).
Because the floats are placed in consecutive banks,
the 1st and 17th threads will access \texttt{float}s from the same bank,
as will the 2nd and 18th threads, and so on.
This specific kind of bank conflict is termed a
\emph{two-way} bank conflict -
threads access two words from the same bank.
It is observed that higher-order bank conflicts may occur,
if threads in a warp access
\texttt{float}s in strides of 4,
a \emph{four-way} bank conflict can occur.
In the worst case, successive threads may access
\texttt{float}s in strides of 32,
in which case all threads in the warp
request memory from the same bank:
a \emph{32-way} bank conflict.

@@

While shared memory and registers can service memory requests
much faster than global memory,
their overuse can lead to performance degradation.
The amount of shared memory per SM
and the number of registers per SM is limited.
For the current NVIDIA Tesla K20 accelerator,
the amount of shared memory per SM is limited to 48 KiB,
and the number of registers per SM is limited to 65536.
These are also the limits
on the resources that can be allocated for each \emph{block}.
However,
allocating 48 KiB shared memory or using 65536 registers 
for each block is ill-advised,
as this effectively restricts the number of blocks that
each SM can run at any given time to 1.
If each block allocates 24 KiB of shared memory,
then an SM can run only 2 blocks concurrently.
Thus, the resources allocated per-block affects
the overall parallelism that can be exploited from the GPU.

@@

    \item The limited amount of shared memory places restrictions
        on the size of tridiagonal systems that can be solved,
        and also on the number of systems that can be solved
        concurrently.
\end{enumerate}

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
fix typos

The fact that the reduced system at each step
is near-Toeplitz can be exploited
to reduce the cost of
precomputing and storing the coefficients.
Each near-Toeplitz matrix is completely defined
by only a handful of coefficients:
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
making its storage extremely compact
compared to the case of general tridiagonal systems.
In addition,
it is advantageous to precompute and store the
auxiliary variables $k_1$ and $k_2$,
which can similarly be stored compactly.
With all the forward reduction coefficient matrices
and auxillary variables precomputed and stored,
the $m$th forward reduction step for equation $i$
is reduced only to the right hand side update:

>>>>

The fact that the reduced system at each step
is near-Toeplitz can be exploited
to reduce the cost of
precomputing and storing the coefficients.
Each near-Toeplitz matrix is completely defined
by only a handful of coefficients:
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
making its storage extremely compact
compared to the case of general tridiagonal systems.
In addition,
it is advantageous to precompute and store the
auxiliary variables $k_1$ and $k_2$,
which can similarly be stored compactly.
With all the forward reduction coefficient matrices
and auxiliary variables precomputed and stored,
the $m$th forward reduction step for equation $i$
is reduced only to the right hand side update:

### 2015-11-18 thesis/results.tex
fix typos

The timings reported are \emph{wall clock} times
with global synchronization between processes performed 
before and after evaluation of the derivatives.
All timings are averaged over 100 evaluations of the function derivatives
in each coordinate direction.
We make it clear that our reported
problem sizes represent the \emph{actual size of problem data}.
Although it may be considered sufficient to
run tests on a single line of processes
for measuring the compact finite difference solver performance,
we set up and solve the problem for the entire computational domain.
In the context of a larger simulation,
global synchronization between the processes
is typically performed before and after the evalaution of derivatives,
and it is important to consider the related overhead.

>>>>

The timings reported are \emph{wall clock} times
with global synchronization between processes performed 
before and after evaluation of the derivatives.
All timings are averaged over 100 evaluations of the function derivatives
in each coordinate direction.
We make it clear that our reported
problem sizes represent the \emph{actual size of problem data}.
Although it may be considered sufficient to
run tests on a single line of processes
for measuring the compact finite difference solver performance,
we set up and solve the problem for the entire computational domain.
In the context of a larger simulation,
global synchronization between the processes
is typically performed before and after the evaluation of derivatives,
and it is important to consider the related overhead.
